---
title: "The tesselle universe"
slug: tesselle-universe
description: |
  Install tesselle packages from our CRAN-like repository on R-universe.
categories:
  - packages
preview: runiverse.png
author:
  - name: Nicolas Frerebeau
    url: https://www.archaeo.science
    affiliation: CREAP (UAR 3414)
    affiliation_url: https://mshs.univ-toulouse.fr
    orcid_id: 0000-0001-5759-4944
date: 2021-08-15
output:
  distill::distill_article:
    toc: true
    toc_float: true
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

Most of the R packages of the **tesselle** project are distributed on [<abbr title="Comprehensive R Archive Network">CRAN</abbr>](https://cran.r-project.org/) when they reach an acceptable level of maturity. Development versions can be installed from [GitHub](https://github.com/tesselle) using [remotes](https://remotes.r-lib.org) :

```{r}
## Install the development version from GitHub
install.packages("remotes")
remotes::install_github("tesselle/tabula")
```

Thanks to the [R-universe](https://ropensci.org/r-universe/) project powered by [rOpenSci](https://ropensci.org), **tesselle** now has its own <abbr title="Comprehensive R Archive Network">CRAN</abbr>-like repository: [**tesselle.r-universe.dev**](https://tesselle.r-universe.dev). This repository allows to install all development versions and non-<abbr title="Comprehensive R Archive Network">CRAN</abbr> packages (e.g. [<abbr title="work-in-progress">WIP</abbr>](https://www.repostatus.org/#wip) packages and large data packages).

## Download tesselle packages

If you want to install a package from the [tesselle universe](https://tesselle.r-universe.dev), simply use `install.packages()` with the additional repository:

```{r}
## Install a package from the 'tesselle' universe
install.packages("tabula", repos = "https://tesselle.r-universe.dev")
```

```{r}
## Enable the repository globally
options(repos = c(tesselle = "https://tesselle.r-universe.dev",
                  CRAN = "https://cloud.r-project.org"))

## Install some packages
install.packages(c("tabula", "kairos"))
```

## Use the repository

You can use this repository in your R package (if available, <abbr title="Comprehensive R Archive Network">CRAN</abbr> releases must be regarded as the preferred source). To do so, prepare the `DESCRIPTION` file of your R package:

* List the package under `Suggests:`
* Add the line `Additional_repositories: https://tesselle.r-universe.dev`
* Test your package with `R CMD check --as-cran`
